---
title: "Create A Package Skeleton"
author: "Peter DeWitt"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteIndexEntry{Create A Package Skeleton}
 %\VignetteEncoding{UTF-8}
---

```{r label=setup, include = FALSE}
library(knitr)
knitr::opts_chunk$set(collapse = TRUE)
```


The `create_pkg`, and its sister functions `create_data_raw` and
`create_vignette`, are provided to build a package skeleton that I prefer.
These functions are based on the
[devtools](https://cran.r-project.org/package=devtools) `create`,
`use_vignette` and `use_data_raw` functions.

This vignette will provided a detailed explanation of the package structure
and design I prefer for R package development.

This vignette only needs the `qwraps2` namespace loaded and attached.

```{r }
library(qwraps2)
```


# Section 1: `create_pkg`

The basic package skeleton is created with the `create_pkg` function.  

```{r }
str(create_pkg)
```


The `path` is a directory to use for the package.  For the examples to follow
we will use a temporary directory.  


```{r }
tmp_dir <- tempdir()
pkg_dir <- paste(tmp_dir, "eg.pkg", sep = "/")
pkg_dir
create_pkg(pkg_dir)

files <- list.files(pkg_dir, all.files = TRUE, full.names = TRUE, recursive = TRUE, include.dirs = TRUE)

data.tree::as.Node(data.frame(filename = files), pathName = "filename")
```


`use_data_raw` sets up the `data` and
`data-raw` directories.  If this



# Section 2



# Session Info


```{r }
print(sessionInfo(), local = FALSE)
```

