---
title: "qwraps2 Summary Table"
author: "Peter E. DeWitt"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: true
vignette: >
 %\VignetteIndexEntry{qwraps2 Summary Table}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---


```{r label = "setup", include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```



```{r}
set.seed(42)
library(qwraps2)
options(qwraps2_markup = "markdown")
```


# Introduction




# Building a Data Summary Table

Objective: build a table reporting summary statistics for some of the variables
in the
`r  backtick(mtcars2)  `
data.frame overall and within subgroups.  We'll start with
something very simple and build up to something bigger.

Let's report the min, max, and mean (sd) for continuous variables and n (%) for
categorical variables.  We will report mpg, displacement (disp), wt (weight),
and gear overall and by number of cylinders and transmission type.

**END USER VISIBLE CHANGE:** for qwraps2 version before 0.4.2 the
`r  backtick(summary_table)  `
method relied on
`r  CRANpkg(dplyr)  `
verbs for the implementation and end user specifications.  This created
several limitations and required what could be considered a non-intuitive api
due to the use of the rlang data pronoun
`r  paste0(backtick(.data), ".")  `
Building a table with the summary based on a grouping, e.g., mpg by number of
cylinders, was achieved by the use of
`r  backtick(dplyr::group_by)  `
to specify the group.  Further only one grouping variable was supported.
Starting with version 0.5.0 the implementation of the
`r  backtick(summary_table)  `
and
`r  backtick(qsummary)  `
is based on base R methods.  The change in the implementation will make it
easier for all users as the use of the tidyverse is no longer required or
assumed.  The use of
`r  backtick(dplyr::group_by)  `
is still supported, and has been improved.

There are two changes to the API for defining summaries:
1. Use of the rlang data pronoun
`r  paste0(backtick(.data), ".")  `
is no longer recommend / supported. Version 0.5.3 a warning will be given
noting that the use of the pronoun _might_ result in unexpected behavior.
2. A new function argument
`r  backtick(by)  `
as been added to the
`r  backtick(summary_table)  `
method such that the use of
`r  backtick(dplyr::group_by)  `
is no longer needed.

The use of the
`r  backtick(summary_table)  `
use to define a summary, that is, a list-of-lists of formulas for summarizing
the data.frame.

The inner lists are named formulae defining the wanted
summary.  The names are important, as they are used to label row groups and row
names in the table.


```{r}
our_summary1 <-
  list("Miles Per Gallon" =
       list("min"       = ~ min(mpg),
            "max"       = ~ max(mpg),
            "mean (sd)" = ~ qwraps2::mean_sd(mpg)),
       "Displacement" =
       list("min"       = ~ min(disp),
            "median"    = ~ median(disp),
            "max"       = ~ max(disp),
            "mean (sd)" = ~ qwraps2::mean_sd(disp)),
       "Weight (1000 lbs)" =
       list("min"       = ~ min(wt),
            "max"       = ~ max(wt),
            "mean (sd)" = ~ qwraps2::mean_sd(wt)),
       "Forward Gears" =
       list("Three" = ~ qwraps2::n_perc0(gear == 3),
            "Four"  = ~ qwraps2::n_perc0(gear == 4),
            "Five"  = ~ qwraps2::n_perc0(gear == 5))
       )
```


Building the table is done with a call to
`r  backtick(summary_table)  `


```{r results = "asis"}
### Overall
whole <- summary_table(mtcars2, our_summary1)
whole

#qable(whole[[1]], rgroup = attr(whole[[1]], "rgroup"), rtitle = "r", cnames = )

### By number of Cylinders
by_cyl <- summary_table(mtcars2, summaries = our_summary1, by = c("cyl_factor"))
by_cyl
```


With the refactor of the
`r  backtick(summary_table)  `
method in version 0.5.0 it is easier to group by multiple variables.  For
example, getting a column for combination of cylinders and transmission type:

```{r}
by_cyl_am <- summary_table(mtcars2, summaries = our_summary1, by = c("cyl_factor", "am"))
by_cyl_am
```


To report a table with both the whole sample summary and conditional columns
together:

```{r results = "asis"}
both <- cbind(whole, by_cyl)
both
```


If you want to change the column names, do so via the
`r  backtick(cnames)  `
argument to
`r  backtick(qable)  `
via the print method for
`r  backtick(qwraps2_summary_table)  `
objects.  Any argument
that you want to send to
`r  backtick(qable)  `
 can be sent there when explicitly using the
`r  backtick(print)  `
method for
`r  backtick(qwraps2_summary_table)  `
objects.

```{r results = "asis"}
print(both,
      rtitle = "Summary Statistics",
      cnames = c("Col 0", "Col 1", "Col 2", "Col 3"))
```


## Easy building of the summaries

The task of building the
`r  backtick(summaries)  `
list-of-lists can be tedious. The function
`r  backtick(qummaries)  `
is designed to make it easier.
`r  backtick(qummaries)  `
will use a set of predefined
functions to summarize numeric columns of a data.frame, a set of arguments
to pass to
`r  backtick(n_perc)  `
for categorical (character and factor) variables.

By default, calling
`r  backtick(summary_table)  `
will use the default summary metrics
defined by
`r  paste0(backtick(qsummary), ".")  `
The purpose of
`r  backtick(qsummary)  `
is to provide the same
summary for all numeric variables within a data.frame and a single style of
summary for categorical variables within the data.frame.  For example, the
default summary for a set of variables from the
`r  backtick(mtcars2)  `
data set is

```{r}
qsummary(mtcars2[, c("mpg", "cyl_factor", "wt")])
```


That default summary is used for a table as follows:

```{r label="summary_table_mtcars2_default", results = "asis"}
summary_table(mtcars2[, c("mpg", "cyl_factor", "wt")])
```


Now, say we want to only report the minimum and maximum for each of the
numeric variables and for the categorical variables we want two show the
denominator for each category and for the percentage, to one digit with the
percent symbol in the table.
Note that when defining the list of numeric_summaries that the argument place
holder is the
`r  backtick("%s%", dequote = TRUE)  `
character.

```{r}
new_summary <-
  qsummary(mtcars2[, c("mpg", "cyl_factor", "wt")],
           numeric_summaries = list("Minimum" = "~ min(%s)",
                                    "Maximum" = "~ max(%s)"),
           n_perc_args = list(digits = 1, show_symbol = TRUE, show_denom = "always"))
str(new_summary)
```


The resulting table is:

```{r results = "asis"}
summary_table(mtcars2, new_summary)
```


The summary can easily be used with a
`r  backtick(by)  `
argument

```{r results = "asis"}
summary_table(mtcars2, new_summary, by = c("cyl_factor"))
```




## Adding P-values to a Summary Table

Starting with qwraps2 version 0.6, the task of adding p-values to a summary
table has been made considerably easier.  This is due to a change in how the
function
`r  backtick(qable)  `
generated the structured matrix.  Older versions of
`r  backtick(qable)  `
generated a structured matrix and passed that matrix to
`r  backtick(knitr::kable)  `
and returned the formatted character string.  In version 0.6
`r  backtick(qable)  `
was changed to return the structured matrix and a separate print method was
added to make the call to
`r  backtick(knitr::kable) %s% "." `

There are many different ways to format data summary tables. Adding
p-values to a table is just one thing that can be done in more than one way.
For example, if a row group reports the counts and percentages for each level
of a categorical variable across multiple (column) groups, then I would argue
that the p-value resulting from a chi square test or a Fisher exact test
would be best placed on the line of the table labeling the row group.
However, say we reported the minimum, median, mean, and maximum with in a
row group for one variable.  The p-value from a t-test, or other meaningful
test for the difference in mean, I would suggest should be reported on the
line of the summary table for the mean, not the row group itself.

With so many possibilities I have reserved construction of a p-value column
to be ad hoc.  Perhaps an additional column wouldn't be used and the p-values
are edited into row group labels, for example.

If you want to add a p-value column to a
`r  backtick(qwraps2_summary_table)  `
object you can with some degree of ease.  Note that
`r  backtick(qwraps2_summary_table)  `
objects are just character matrices.

```{r}
str(both)
```


Let's added p-values for testing the difference in the mean between the three
cylinder groups.

```{r}
# difference in means
mpvals <-
  sapply(
         list(lm(mpg ~ cyl_factor,  data = mtcars2),
              lm(disp ~ cyl_factor, data = mtcars2),
              lm(wt ~ cyl_factor,   data = mtcars2)),
         extract_fpvalue)

# Fisher test
fpval <- frmtp(fisher.test(table(mtcars2$gear, mtcars2$cyl_factor))$p.value)
```


Adding the p-value column is done as follows:

```{r}
both <- cbind(both, "P-value" = "")
both[grepl("mean \\(sd\\)", both[, 1]), "P-value"] <- mpvals
both[grepl("Forward Gears", both[, 1]), "P-value"] <- fpval
```


and the resulting table is:

```{r results = "asis"}
both
```


Another option you might consider is to have the p-value in the row group
name.  Consider the following construction.  The p-values are added to the
names of the row groups when building the summary table.

```{r results = "asis"}
gear_summary <-
  list("Forward Gears" =
       list("Three" = ~ qwraps2::n_perc0(gear == 3),
            "Four"  = ~ qwraps2::n_perc0(gear == 4),
            "Five"  = ~ qwraps2::n_perc0(gear == 5)),
       "Transmission" =
       list("Automatic" = ~ qwraps2::n_perc0(am == 0),
            "Manual"    = ~ qwraps2::n_perc0(am == 1))
       )

gear_summary <-
setNames(gear_summary,
         c(
         paste("Forward Gears: ", frmtp(fisher.test(xtabs( ~ gear + cyl_factor, data = mtcars2))$p.value)),
         paste("Transmission: ",  frmtp(fisher.test(xtabs( ~ am + cyl_factor, data = mtcars2))$p.value)))
         )

summary_table(mtcars2, gear_summary, by = "cyl_factor")
```


## Using Variable Labels

Some data management paradigms will use attributes to keep a label associated
with a variable in a data.frame.  Notable examples are the
`r  CRANpkg(Hmisc)  `
and
`r  paste0(CRANpkg(sjPlot), ".")  `
If you associate a label with a variable in the data frame the that label
will be used when building a summary table.  This feature was suggested
https://github.com/dewittpe/qwraps2/issues/74 and implemented thusly:

```{r}
new_data_frame <-
  data.frame(age = c(18, 20, 24, 17, 43),
             edu = c(1, 3, 1, 5, 2),
             rt  = c(0.01, 0.04, 0.02, 0.10, 0.06))

# Set a label for the variables
attr(new_data_frame$age, "label") <- "Age in years"
attr(new_data_frame$rt,  "label") <- "Reaction time"

# mistakenly set the attribute to name instead of label
attr(new_data_frame$edu, "name") <- "Education"
```


When calling
`r  backtick(qsummary)  `
the provide labels for the age and rt variables will
be used.  Since the attribute "label" does not exist for the edu variable,
edu will be used in the output.

```{r}
qsummary(new_data_frame)
```


This behavior is also seen with the
`r  backtick(summary_table)  `
call.

```{r results = "asis"}
summary_table(new_data_frame)
```



# Session Info


```{r}
print(sessionInfo(), local = FALSE)
```

