---
title: "Formatted Summary Statistics and Data Summary Tables with qwraps2"
author: "Peter DeWitt"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{summary-statistics}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---


```{r label = "setup", include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```



# Introduction

It is common for a manuscript to require a data summary table.  The table might
include simple summary statistics for the whole sample and for subgroups.
There are several tools available to build such tables.  In my opinion, though,
most of those tools have nuances imposed by the creators/authors such that other
users need not only understand the tool, but also think like the authors.
I wrote this package to be as flexible and general as possible.  I hope you like
these tools and will be able to use them in your work.

This vignette presents the use of the `summary_table`, `tab_summary`, and
`qable` functions for quickly building data summary tables.  These functions
implicitly use the `mean_sd`, `median_iqr`, and `n_perc0` functions from
`qwraps2` as well.

## Prerequisites Example Data Set
We will use a modified version of the `mtcars` data set for examples throughout
this vignette. The following packages are required to run the code in this
vignette and to construct the `mtcars2` `data.frame`.

The `mtcars2` data frame will have three versions of the `cyl` vector: the
original numeric values in `cyl`, a `character` version, and a `factor` version.

```{r }
set.seed(42)
library(dplyr)
library(qwraps2)

# define the markup language we are working in.
# options(qwraps2_markup = "latex") is also supported.
options(qwraps2_markup = "markdown")

data(mtcars)

mtcars2 <-
  dplyr::mutate(mtcars,
                cyl_factor = factor(cyl,
                                    levels = c(6, 4, 8),
                                    labels = paste(c(6, 4, 8), "cylinders")),
                cyl_character = paste(cyl, "cylinders"))

str(mtcars2)
```


Notice that the construction of the `cyl_factor` and `cyl_character` vectors
was done such that the coercion of `cyl_character` to a `factor` will not be the
same as the `cyl_factor` vector; the levels are in a different order.


```{r }
with(mtcars2, table(cyl_factor, cyl_character))
with(mtcars2, all.equal(factor(cyl_character), cyl_factor))
```


# Review of Summary Statistic Functions and Formatting

## Means and Standard Deviations
`mean_sd` will return the (arithmetic) mean and standard deviation for numeric
vector. For example, `mean_sd(mtcars2$mpg)` will return the formatted string.


```{r }
mean_sd(mtcars2$mpg)
mean_sd(mtcars2$mpg, denote_sd = "paren")
```


The default setting for `mean_sd` is to return the mean &plusmn; sd.  In a
table this default is helpful because the default table formatting for counts
and percentages is n (%).

`mean_sd` and other functions are helpful for in-line text too:

    The `r nrow(mtcars2)` vehicles in the `mtcars` data set had an average fuel
    economy of `r mean_sd(mtcars$mpg)` miles per gallon.

produces

> The `r nrow(mtcars2)` vehicles in the `mtcars` data set had an average fuel
> economy of `r mean_sd(mtcars$mpg)` miles per gallon.

## Mean and Confidence intervals
If you need the mean and a confidence interval there is `mean_ci`.
`mean_ci` returns a `qwraps2_mean_ci` object which is a
named vector with the mean, lower confidence limit, and the upper confidence
limit.   The printing method for `qwraps2_mean_ci` objects is a call to the
`frmtci` function.  You an modify the formatting of printed result by adjusting
the arguments pasted to `frmtci`.

```{r }
mci <- mean_ci(mtcars2$mpg)
mci
print(mci, show_level = TRUE)
```


## Median and Inner Quartile Range
Similar to the `mean_sd` function, the `median_iqr` returns the median and the
inner quartile range (IQR) of a data vector.

```{r }
median_iqr(mtcars2$mpg)
```


## Count and Percentages
The `n_perc` function is the workhorse, but `n_perc0` is also provided for ease
of use in the same way that base R has `paste` and `paste0`.  `n_perc` returns
the n (%) with the percentage sign in the string, `n_perc0` omits the
percentage sign from the string.  The latter is good for tables, the former for
in-line text.


```{r }
n_perc(mtcars2$cyl == 4)
n_perc0(mtcars2$cyl == 4)

n_perc(mtcars2$cyl_factor == 4)  # this returns 0 (0.00%)
n_perc(mtcars2$cyl_factor == "4 cylinders")
n_perc(mtcars2$cyl_factor == levels(mtcars2$cyl_factor)[2])

# The count and percentage of 4 or 6 cylinders vehicles in the data set is
n_perc(mtcars2$cyl %in% c(4, 6))
```


## Geometric Means and Standard Deviations

Let $\left\{x_1, x_2, x_3, \ldots, x_n \right\}$ be a sample of size $n$ with
$x_i > 0$ for all $i.$  Then the geometric mean, $\mu_g,$ and geometric standard
deviation are in Equation \@ref(eq:geometricmean) and \@ref(eq:geometricsd)
respectively.

$$
\begin{equation}
  (\#eq:geometricmean)
  \mu_g = \left( \prod_{i = 1}^{n} x_i \right)^{\frac{1}{n}} = b^{ \sum_{i = 1}^{n} \log_{b} x_i }
\end{equation}
$$

$$
\begin{equation}
  (\#eq:geometricsd)
  \sigma_g = b ^ {
  \sqrt{ \frac{\sum_{i = 1}^{n} \left( \log_{b} \frac{x_i}{\mu_g}
  \right)^2}{n}}}
\end{equation}
$$

When looking for the geometric standard deviation in R, the simple
`exp(sd(log(x)))` is not exactly correct.  Note that in
\@ref(eq:geometricsd) the denominator is $n,$ the full sample size, where as
the `sd` and `var` functions in R use the denominator $n - 1.$  To get
the geometric standard deviation one should adjust the result by multiplying the
variance by $(n - 1) / n$ or the standard deviation by $\sqrt{(n - 1) / n}.$
See the example below.

```{r }
x <- runif(6, min = 4, max = 70)

# geometric mean
mu_g <- prod(x) ** (1 / length(x))
mu_g
exp(mean(log(x)))
1.2 ** mean(log(x, base = 1.2))

# geometric standard deviation
exp(sd(log(x)))  ## This is wrong

# these equations are correct
sigma_g <- exp(sqrt(sum(log(x / mu_g) ** 2) / length(x)))
sigma_g

exp(sqrt((length(x) - 1) / length(x)) * sd(log(x)))
```


The functions `gmean`, `gvar`, and `gsd` in the
package, provide the geometric
mean, variance, and standard deviation for a sample.

```{r }
gmean(x)
all.equal(gmean(x), mu_g)

gvar(x)
all.equal(gvar(x), sigma_g^2)  # This is supposed to be FALSE
all.equal(gvar(x), exp(log(sigma_g)^2))

gsd(x)
all.equal(gsd(x), sigma_g)
```


`gmean_sd` will provide a quick way for reporting the geometric mean and
geometric standard deviation in the same way that `mean_sd` does for the
arithmetic mean and arithmetic standard deviation:

```{r }
gmean_sd(x)
```


# Building a Data Summary Table

Objective: build a table reporting summary statistics for some of the variables
in the `mtcars2` `data.frame` overall and within subgroups.  We'll start with
something very simple and build up to something bigger.

Let's report the min, max, and mean (sd) for continuous variables and n (%) for
categorical variables.  We will report `mpg`, `disp`, `wt`, and `gear` overall
and by number of cylinders.

The function `summary_table`, along with some `dplyr` functions will do the work
for us.  `summary_table` takes two arguments:

1. `.data` a (`grouped_df`) data.frame
2. `summaries` a list of summaries.  This is a list-of-lists.  The outer list
   defines the row groups and the inner lists define the specif summaries.


```{r }
args(summary_table)
```


Let's build a list-of-lists to pass to the `summaries` argument of
`summary_table`.  The inner lists are named `formula`e defining the wanted
summary.  These `formula`e are passed through `dplyr::summarize_` to generate
the table.  The names are important, as they are used to label row groups and row
names in the table.


```{r }
our_summary1 <-
  list("Miles Per Gallon" =
       list("min" = ~ min(mpg),
            "max" = ~ max(mpg),
            "mean (sd)" = ~ qwraps2::mean_sd(mpg)),
       "Displacement" =
       list("min" = ~ min(disp),
            "max" = ~ max(disp),
            "mean (sd)" = ~ qwraps2::mean_sd(disp)),
       "Weight (1000 lbs)" =
       list("min" = ~ min(wt),
            "max" = ~ max(wt),
            "mean (sd)" = ~ qwraps2::mean_sd(wt)),
       "Forward Gears" =
       list("Three" = ~ qwraps2::n_perc0(gear == 3),
            "Four"  = ~ qwraps2::n_perc0(gear == 4),
            "Five"  = ~ qwraps2::n_perc0(gear == 5))
       )
```


Building the table is done with a call to `summary_table`:


```{r results = "asis"}
### Overall
summary_table(mtcars2, our_summary1)
summary_table(mtcars2, our_summary1)
```





```{r results = "asis"}
### By number of Cylinders
summary_table(dplyr::group_by(mtcars2, cyl_factor), our_summary1)
```


If you want to change the column names, do so via the `cnames` argument to
`qable` via the print method for `qwraps2_summary_table` objects.  Any argument
that you want to send to `qable` can be sent there when explicitly using the
`print` method for `qwraps2_summary_table` objects.

```{r results = "asis"}
print(summary_table(dplyr::group_by(mtcars2, cyl_factor), our_summary1),
      rtitle = "Summary Statistics",
      cnames = c("Col 1", "Col 2", "Col 3"))
```


## Easy building of the summaries
The task of building the `summaries` list-of-lists can be tedious.  `tab_summary`
is designed to make it easier.  For `numeric` variables, `tab_summary` will
provide the `formula`e for the min, median (iqr), mean (sd), and max.  `factor`
and `character` vectors will have calls to `qwraps2::n_perc` for all levels
provided.

For version 0.2.3.9000 or beyond, arguments have been added to `tab_summary` to
help control some of the formatting of counts and percentages.  The original
behavior of `tab_summary` used `n_perc0` to format the summary of categorical
variables.  Now, `n_perc` is called and the end user can specify formatting
options via a `list` passed via the `n_perc_args` argument.
The default settings for `tab_summary` is below.

```{r }
args(tab_summary)
```


These options will make the output look as if `n_perc0` had been called instead
of `n_perc`.  More importantly, these defaults *will not* honor the
`options()$qwraps2_frmt_digits`.

Examples for `tab_summary` follow:

```{r }
tab_summary(mtcars2$mpg)

tab_summary(mtcars2$gear) # gear is a numeric vector!
tab_summary(factor(mtcars2$gear))
```


The `our_summary1` object can be recreated as follows.  Some additional row
groups are provided to show default behavior of `tab_summary`. **Important:**
Note that the `tab_summary` are made while using `with`.  Further explanation
for this follows.


```{r }
our_summary2 <-
  with(mtcars2,
       list("Miles Per Gallon" = tab_summary(mpg)[c(1, 4, 3)],
            "Displacement (default summary)" = tab_summary(disp),
            "Displacement" = c(tab_summary(disp)[c(1, 4, 3)],
                               "mean (95% CI)" = ~ frmtci(qwraps2::mean_ci(disp))),
            "Weight (1000 lbs)" = tab_summary(wt)[c(1, 4, 3)],
            "Forward Gears" = tab_summary(as.character(gear))
            ))
```





```{r results = "asis"}
whole <- summary_table(mtcars2, our_summary2)
whole
```


Group by multiple factors:


```{r results = "asis"}
grouped <- summary_table(dplyr::group_by(mtcars2, am, vs), our_summary2)
grouped
```


As one table:


```{r results = "asis"}
both <- cbind(whole, grouped)
both
```


### Adding P-values to a Summary Table

There are many, many different ways to format data summary tables. Adding
p-values to a table is just one thing that can be done in more than one way.
For example, if a row group reports the counts and percentages for each level
of a categorical variable across multiple (column) groups, then I would argue
that the p-value resulting from a chi square test or a Fisher exact test
would be best placed on the line of the table labeling the row group.
However, say we reported the minimum, median, mean, and maximum with in a
row group for one variable.  The p-value from a t-test, or other meaningful
test, for the difference in mean I would suggest should be reported on the
line of the summary table for the mean, not the row group itself.

With so many possibilities I have reserved construction of a p-value column
to be ad hoc.  Perhaps an additional column wouldn't be used and the p-values
are edited into row group labels, for example.

If you want to add a p-value column to a `qwraps2_summary_table` object you
can with some degree of ease.  Note that `qwraps2_summary_table` objects are
just character matrices.

```{r }
both %>% str

# another good way to veiw the character matrix
# print.default(both)
```


Let's added p-values for testing the difference in the mean between the four
groups defined by `am:vs`.

```{r }
pvals <-
  list(lm(mpg ~ am:vs,  data = mtcars2),
       lm(disp ~ am:vs, data = mtcars2),
       lm(disp ~ am:vs, data = mtcars2),  # yeah, silly example this is needed twice
       lm(wt ~ am:vs,   data = mtcars2)) %>%
  lapply(aov) %>%
  lapply(summary) %>%
  lapply(function(x) x[[1]][["Pr(>F)"]][1]) %>%
  lapply(frmtp) %>%
  do.call(c, .)
pvals
```


Adding the p-value column is done as follows:

```{r }
both <- cbind(both, "P-value" = "")
both[grepl("mean \\(sd\\)", rownames(both)), "P-value"] <- pvals
```


and the resulting table is:

```{r results = "asis"}
both
```


### Why use `with` with `tab_summary`?
`tab_summary` was written to help construct `formula`e to save the end user key
strokes.  There are plenty of reasons for `summary_table` to be used *without*
`tab_summary`.  However, when it is helpful to use `tab_summary` make sure you
understand the results.

For example, let's look at a simple summary of the miles per gallon.

```{r }
# tab_summary(mpg) ## this errors
tab_summary(mtcars$mpg)
with(mtcars, tab_summary(mpg))
```


The first call errors because `mpg` is not in the global environment.  The
difference between the second and third calls is subtle.  The second call
generates a `formula` with `mtcars$mpg` as an argument whereas the third call
generates a `formula` with only `mpg` as the argument.  The difference will be
seen in the summary tables if the `.data` is subsetted.

```{r }
# The same tables:
summary_table(mtcars, list("MPG 1" = with(mtcars, tab_summary(mpg))))
summary_table(mtcars, list("MPG 2" = tab_summary(mtcars$mpg)))
```


These two calls generate the same table because the `.data` and the implied
data within the second call are the same.

```{r results = "asis"}
# Different tables
summary_table(dplyr::filter(mtcars, am == 0), list("MPG 3" = with(mtcars, tab_summary(mpg))))
summary_table(dplyr::filter(mtcars, am == 0), list("MPG 4" = tab_summary(mtcars$mpg)))
```


Now, the result of the second call above is not correct, it is the same as
for the first two calls.  This is because `mtcars$` is part of the `formula`
and the `.data` is ignored.  The correct result is in the table with `MPG 3`.

I encourage you, the end user, to use `summary_table` primarily, and use
`tab_summary` as a quick tool for generating a script.  It might be best if
you use `tab_summary` to generate a template of the `formula`e you will want,
copy the template into your script and edit accordingly.

# Session Info

```{r }
print(sessionInfo(), local = FALSE)
```

