---
title: "Formatted Summary Statistics and Data Summary Tables with qwraps2"
author: "Peter DeWitt"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{summary-statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE) 
```

# Introduction

It is common for a manuscript to require a data summary table.  Simple summary
statistics for the whole sample and for subgroups therein.  There are many tools
available for easing the difficulty involved in building such tables.   It is my
opinion, that most of these tool exists with sufficient implicit biases and
nuances imposed by the authors of the tools that other users need to not only
understand the tool, but to think like the author.  I find myself approaching
many problems in ways that few others do.  As a result, I needed my own tool for
building data summary tables.  I hope you like these tools and will be able to
use it in your work. 

This vignette presents the use of the `summary_table`, `tab_summary`, and
`qable` functions for quickly building data summary tables.  These functions
implicitly use the `mean_sd`, `median_iqr`, and `n_perc0` functions from
`qwraps2` as well.

## Prerequisites Example Data Set
We will use a modified version of the `mtcars` data set for examples throughout
this vignette. The following packages are required to run the code in this
vignette and to construct the `mtcars2` `data.frame`.

The `mtcars2` data frame will have three versions of the `cyl` vector, the
original numeric values in `cyl`, a `character` version, and a `factor` version.

```{r}
library(dplyr)
library(qwraps2)

# define the markup language we are working in.
# options(qwraps2_markup = "latex") is also supported.
options(qwraps2_markup = "markdown")

data(mtcars)

mtcars2 <- 
  dplyr::mutate(mtcars,
                cyl_factor = factor(cyl, 
                                    levels = c(6, 4, 8), 
                                    labels = paste(c(6, 4, 8), "cylinders")),
                cyl_character = paste(cyl, "cylinders"))

str(mtcars2)
```
Notice that in the construction of the `cyl_factor` and `cyl_character` vectors
was done such that the coercion of `cyl_character` to a `factor` will not be the
same as the `cyl_factor` vector, the levels are in a different order.

```{r}
with(mtcars2, table(cyl_factor, cyl_character))
with(mtcars2, all.equal(factor(cyl_character), cyl_factor))
```

# Review of Summary Statistic Functions and Formatting
Let's review some of the formatting functions provided by the `qwraps2` package

## Means and Standard Deviations
`mean_sd` will return the (arithmetic) mean and standard deviation for numeric
vector, i.e., `mean_sd(mtcars2$mpg)` will return the formatted string.
```{r}
mean_sd(mtcars2$mpg)
mean_sd(mtcars2$mpg, denote_sd = "paren") 
```
The default setting for `mean_sd` is to return the mean &plusmn; sd.  In a
table this default is helpful sense the default formating for counts and
percentages is n (%).  

The `mean_sd`, and other functions, are helpful for in-line text too:

> The `nrow(mtcars2)` vehicles in the `mtcars` data set had an average fuel
> economy of `mean_sd(mtcars$mpg) miles per gallon.

produces

> The `r nrow(mtcars2)` vehicles in the `mtcars` data set had an average fuel
> economy of `r mean_sd(mtcars$mpg)` miles per gallon.

## Mean and Confidence intervals
If you need the mean and a confidence interval there is `mean_ci`.
`mean_ci` returns a `qwraps2_mean_ci` object which is a
named vector with the mean, lower confidence limit, and the upper confidence
limit.   There printing method for `qwraps2_mean_ci` objects is a call to the
`frmtci` function.  You an modify the formating of printed result via adjusting
the arguments pasted to `frmtci`
```{r}
mci <- mean_ci(mtcars2$mpg)
mci
print(mci, show_level = TRUE)
```

## Median and Inner Quartile Range
Similar to the `mean_sd` function, the `median_iqr` returns the median and the
inner quartile range (IQR) of a data vector.
```{r}
median_iqr(mtcars2$mpg)
```

## Count and Percentages
The `n_perc` function is the workhorse, but `n_perc0` is also provided for ease
of use in the same way that base R has `paste` and `paste0`.  `n_perc` returns
the n (%) with the percentage sign in the string, `n_perc0` omits the
percentage sign from the string.  The latter is good for table, the former for
in-line text.

```{r}
n_perc(mtcars2$cyl == 4)
n_perc0(mtcars2$cyl == 4)

n_perc(mtcars2$cyl_factor == 4)  # this returns 0 (0.00%) 
n_perc(mtcars2$cyl_factor == "4 cylinders")
n_perc(mtcars2$cyl_factor == levels(mtcars2$cyl_factor)[2]) 

# The count and percentage of 4 or 6 cylinders vehicles in the data set is
n_perc(mtcars2$cyl %in% c(4, 6))
```

# Building a Data Summary Table

Objective: build a table reporting summary statistics for some of the variables
in the `mtcars2` `data.frame` overall, and within subgroups.  We'll start with
something very simple and build up to something bigger.

Let's report the min, max, and mean (sd) for continuous variables and n (%) for
categorical variables.  We will report `mpg`, `disp`, `wt`, and `gear` overall
and by number of cylinders.

The function `summary_table`, along with some `dplyr` functions will do the work
for us.  `summary_table` takes two arguments:

1. `.data` a (`grouped_df`) data.frame
2. `summaries` a list of summaries.  This is a list-of-lists.  The outer list
   defines the row groups and the inner lists define the specif summaries.

```{r}
args(summary_table)
```

Let's build a list-of-lists to pass to the `summaries` argument of
`summary_table`.  The inner lists are named `formula`e defining the wanted
summary.  These `formula`e are passed through `dplyr::summarize_` to generate
the table.  The names are important, as they are used to label row groups and row
names in the table.

```{r}
our_summary1 <- 
  list("Miles Per Gallon" = 
       list("min" = ~ min(mpg), 
            "max" = ~ max(mpg),
            "mean (sd)" = ~ qwraps2::mean_sd(mpg)),
       "Displacement" = 
       list("min" = ~ min(disp), 
            "max" = ~ max(disp),
            "mean (sd)" = ~ qwraps2::mean_sd(disp)),
       "Weight (1000 lbs)" = 
       list("min" = ~ min(wt), 
            "max" = ~ max(wt),
            "mean (sd)" = ~ qwraps2::mean_sd(wt)),
       "Forward Gears" = 
       list("Three" = ~ qwraps2::n_perc0(gear == 3),
            "Four"  = ~ qwraps2::n_perc0(gear == 4),
            "Five"  = ~ qwraps2::n_perc0(gear == 5))
       ) 
```

Building the table is done with a call to `summary_table`:

### Overall
```{r, results = "asis"}
summary_table(mtcars2, our_summary1)
summary_table(mtcars2, our_summary1)
```

### By number of Cylinders
```{r, results = "asis"}
summary_table(dplyr::group_by(mtcars2, cyl_factor), our_summary1)
```

If you want to change the column names do so via the `cnames` argument to
`qable` via the print method for `qwraps2_summary_table` objects.  Any argument
that you want to send to `qable` can be sent there when explicitly using the
`print` method for `qwraps2_summary_table` objects.
```{r, results = "asis"}
print(summary_table(dplyr::group_by(mtcars2, cyl_factor), our_summary1), 
      rtitle = "Summary Statistics",
      cnames = c("Col 1", "Col 2", "Col 3"))
```

## Easy building of the summaries
The task building the `summaries` list-of-lists can be tedious.  `tab_summary`
is provided to help with that.  For `numeric` variable, `tab_summary` will
provide the `formula`e for the min, median (iqr), mean (sd), and max.  `factor`
and `character` vectors will have calls to `qwraps2::n_perc0` for all levels
provided.

```{r}
tab_summary(mtcars2$mpg)

tab_summary(mtcars2$gear) # gear is a numeric vector!
tab_summary(factor(mtcars2$gear)) 
```

The `our_summary1` object can be recreated as follows.  Some additional row
groups are provided to show default behavior of `tab_summary`.

```{r}
our_summary2 <- 
  with(mtcars2, 
       list("Miles Per Gallon" = tab_summary(mpg)[c(1, 4, 3)],
            "Displacement (default summary)" = tab_summary(disp),
            "Displacement" = c(tab_summary(disp)[c(1, 4, 3)],
                               "mean (95% CI)" = ~ frmtci(qwraps2::mean_ci(disp))),
            "Weight (1000 lbs)" = tab_summary(wt)[c(1, 4, 3)],
            "Forward Gears" = tab_summary(as.character(gear))
            ))
```

```{r, results = "asis"}
whole <- summary_table(mtcars2, our_summary2)
whole
```

Group by multiple factors:
```{r, results = "asis"}
grouped <- summary_table(dplyr::group_by(mtcars2, am, vs),  our_summary2)
grouped
```

As one table:

```{r, results = "asis"}
both <- cbind(whole, grouped)
both
```


