---
title: "summary table"
output: rmarkdown::word_document
---

To get the summary table from the
`r  qwraps2::CRANpkg(qwraps2) `
package to render in a word document you need to *not* pass the return from
`r  qwraps2::backtick(summary_test) `
to
`r  paste0(qwraps2::backtick(knitr::kabel), ".") `

The reason is that
`r  qwraps2::backtick(summary_test) `
returns a character matrix and a
`r  qwraps2::backtick(print) `
method is reasponsible for generating either the markdown or LaTeX table.


```{r label = "install_cran", include = FALSE}
# remotes::install_cran is similar to your check_and_install function.
remotes::install_cran("kableExtra")
remotes::install_cran("dplyr")
remotes::install_cran("qwraps2")

```
```{r label = "namespaces"}
library(kableExtra)
library(qwraps2)
```


By default
`r  qwraps2::CRANpkg(qwraps2) `
will format results for LaTeX.  Set the default markup language to markdown
to change this behavior.

```{r }
options(qwraps2_markup = "markdown")
```

**EDIT:** as of version qwraps2 version 0.5.0 the use of the data pronoun
`.data` is no longer needed nor recommented.

```{r label = "define_summary_test"}
summary_test  <-
  list("Cylindres" =
     list("Huit"   = ~ qwraps2::n_perc0(cyl == 8, show_symbol = TRUE),
          "Six"    = ~ qwraps2::n_perc0(cyl == 6, show_symbol = TRUE),
          "Quatre" = ~ qwraps2::n_perc0(cyl == 4, show_symbol = TRUE)),
   "Vitesses" =
     list("Cinq"   = ~ qwraps2::n_perc0(gear == 5, show_symbol = TRUE),
          "Quatre" = ~ qwraps2::n_perc0(gear == 4, show_symbol = TRUE),
          "Trois"  = ~ qwraps2::n_perc0(gear == 3, show_symbol = TRUE))
  )
```


Finally, build the table and look at the structure of the object.


```{r label = 'tabtest2'}
tabtest2 <- summary_table(mtcars, summary_test, by = "am")
str(tabtest2)
```


Note that the object is a
`r  qwraps2::backtick(qwraps2_summary_table)  `
is a character matrix with attributes for the row, column, and row group
names.  Using the default print method in R we see a character matrix.

```{r }
print.default(tabtest2)
```


Using the print method for
`r  qwraps2::backtick(qwraps2_summary_table)  `
objects (done impliclity here) gives the markdown:

```{r results = "markup"}
tabtest2
```


To get the table to render nicely use the "asis" value for the results chunk
option:

```{r results = "asis"}
tabtest2
```
