The code in Chunk 1 compiles, but no caption when the markup language is
markdown.  The table is not a floating environment.

```{r label = "example_table_1", results = "asis"}
# Chunk 1
library(qwraps2)
options(qwraps2_markup = "markdown")
eg_table <- summary_table(mtcars[c("mpg", "hp")])
print(eg_table, caption = "Example Caption")
```

Chunk 2 will work and I can reference the captioned Table
\ref{tab:example_table_2} as I would any other table in \LaTeX.


```{r label = "example_table_2", results = "asis"}
# Chunk 2
options(qwraps2_markup = "latex")
eg_table <- summary_table(mtcars[c("mpg", "hp")])

# use booktabs = TRUE to have better formatting.
print(eg_table, caption = "Example Caption", booktabs = TRUE)
```

The last chunk here, chunk 3 will result in a compile error.  The issue is that
th summary statistics characters are generated with the expectation of markdown
but the markup language wrapping up the character matrix is \LaTeX.  The
\&plusmn; html code introduces the ``Extra alignment tab has been change
to \cr.` error.  

```{r label = "example_table_3", results = "markup"}
# Chunk 3 ... compile error if chunk option "results = 'asis'" is used.  No
options(qwraps2_markup = "markdown")
eg_table <- summary_table(mtcars[c("mpg", "hp")])
print(eg_table, caption = "Example Caption", markup = "latex")

# omit the \hline after every row
print(eg_table, caption = "Example Caption", markup = "latex", booktabs = TRUE)
```


