---
title: "Quick Package Skeletons"
author: "Peter DeWitt"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteIndexEntry{Quick Package Skeletons}
 %\VignetteEncoding{UTF-8}
---

```{r label=setup, include = FALSE}
library(knitr)
knitr::opts_chunk$set(collapse = TRUE)
```


Building R packages has become easier and easier over the last several years.
Two packages,
[devtools](https://github.com/r-libs/devtools) and
[usethis](https://github.com/r-lib/usethis), along with the imported and
suggested packages used by these two packages, are, in my opinion, have made
package authoring a viable option for novice and expert R programmers alike.

Of course every programmer has their own style and I am no exception.  The
qwraps2 package provides the `qpkg` (Quick Package) function to build an R
package skeleton.  This vignette serves as an example for the use of the
function and an explanation of how I prefer to write my R packages.

A couple things to be aware of as you read this vignette:

* I write a lot of packages for data analysis projects.  That is, I'm the
consulting statistician on a research project and I use the R package
structure to document and check my work.  Obviously, functions written to
automate repeated analysis tasks are easy to document in a package.
Vignettes are used to document data-import, exploratory data analysis, and
are the basis for data analysis reports and manuscripts.

* I prefer to work in a Linux environment, Debian is my preferred
distribution, and my preferred IDE is [neovim](https://neovim.io) with the
[Nvim-R](https://github.com/jalvesaq/Nvim-R) plugin. If you are working in
Linux or on a Mac I would expect that the following work will be relatively
easy to follow and replicate.  If you are working on Windows, you might have
some issue arise as the following rely on symbolic links.  Window users will
need:

  * Vista or later with a NTFS file system, not FAT,
  * You need administrator rights or at least `SeCreateSymbolicLinkPrivilege`
  privilege
  * git bash version 2.10.2 or later.  It will be helpful to install with
  `core.symlinks` option turned on.




# Examples:

The best example for this style of package development is the
[qwraps2](https://github.com/dewittpe/qwraps2) package on github.



# Session Info


```{r }
print(sessionInfo(), local = FALSE)
```

