---
title: "Create A Package Skeleton"
author: "Peter DeWitt"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteIndexEntry{Create A Package Skeleton}
 %\VignetteEncoding{UTF-8}
---

```{r label=setup, include = FALSE}
library(knitr)
knitr::opts_chunk$set(collapse = TRUE)
```


The `create_pkg`, and its sister functions `create_data_raw` and
`create_vignette`, are provided to build a package skeleton that I prefer.
These functions are based on the
[devtools](https://cran.r-project.org/package=devtools) `create`,
`use_vignette` and `use_data_raw` functions.

This vignette will provided a detailed explanation of the package structure
and design I prefer for R package development.

This vignette only needs the `qwraps2` namespace loaded and attached.

```{r }
library(qwraps2)
```


# Section 1: `create_pkg`

The basic package skeleton is created with the `create_pkg` function.

```{r }
str(create_pkg)
```


The `path` is a directory to use for the package.  For the examples to follow
we will use a temporary directory.


```{r }
tmp_dir <- tempdir()
pkg_dir <- paste(tmp_dir, "eg.pkg", sep = "/")
pkg_dir
create_pkg(pkg_dir)
```


Let's look at the files and the skeleton created for the package.  We'll do
this a few times so a `tree` function will be created and used:

```{r }
tree <- function(pkg_dir) {
  files <- list.files(pkg_dir,
                      all.files = TRUE,
                      full.names = TRUE,
                      recursive = TRUE,
                      include.dirs = TRUE)
  files <- data.frame(filename = gsub(tmp_dir, "", files), file.info(files)) 
  print(data.tree::as.Node(files, pathName = "filename"), "isdir")
}
tree(pkg_dir)
```


There are two empty directories created, the `R` directory and the `examples`
directory.  The `examples` directory is expected to be used to store the
examples for the documentation written, using
[roxygen2](https://cran.r-project.org/package=roxygen2) markup, in the files
within the `R` directory.  More details are given in a later section.

The files populating the root directory of the package includes the needed
`DESCRIPTION` file, which you, the package developer, will need to edit.
Details on the metadata and what/how to edit it is found in the [R
packages](http://r-pkgs.had.co.nz/description.html) book by Hadley Wickham.

```{r }
cat(readLines(paste0(pkg_dir, "/DESCRIPTION")), sep = "\n")
```


The
`NAMESPACE` file is provided, but **should not** be edited by the end user,
let `devtools::document()`, called via the `makefile` handle the edits to the
`NAMESPACE` file.  General package documentation in the form of a
`README.md`, `CONTRIBUTING.md`, and `NEWS.md` are provided.  The
`.Rbuildignore` and `.gitignore` files are provided with commonly needed
expressions.  Lastly, the `makefile` is provided for building, checking, and
installing the package.  A `help` recipe has been provided in the makefile.

```
make help
```

The recipes in the makefile are:

```{r }
# cat(readLines(paste0(pkg_dir, "/makefile")), sep = "\n")
```

This `makefile` is built to be useful for most R package development, it is
copied into new R packages created by `qwraps2::create_pkg`, see
`vignette("create_pkg", package = "qwraps2")`.

   usage: make [build-options] [check-options] [install-options] [targets]

Target          Description
------          -----------
all             Build the man files, vignettes, and the `<package>_<version>.tar.gz` file
help            Display usage, targets, and options provided by the makefile
check           Run `R CMD check`
install         Install the R package via `R CMD INSTALL`
document        Construct raw data sets, if needed, and build/update man files.
vignettes 

Options       Description
-------       -----------
build-options

If the `rstudio` option is used when calling create_pkg then a .Rproj file is
added to the root directory.

The `ci` argument is used to add template files for either [Travis
CI](https://travis-ci.org) or gitlab runners.

The `use_data_raw` option is used if raw data is to be part of the package.
calling the `create_data_raw` function can be called later to create the
directories and place the generic makefiles as needed to work with the
makefile in the package root directory.

## `create_data_raw`

Adding the `data-raw` and `data` directories to the package, if not done on
the initial construction, is done via

```{r }
create_data_raw(pkg_dir)
```


The updated package structure is:

```{r }
tree(pkg_dir)
```


The makefile in the `data-raw` is generic such that any `.R` file will be
evaluated.  Noting that some scripts needed to create a data set might
require a non-trivial amount of time to evaluate, the makefile creates, and
then uses, md5sums for assessing if the `.R` file(s) needs to be re-evaluated
instead of the standard file modified time stamp.

```{r }
cat(readLines(paste0(pkg_dir, "/data-raw/makefile")), sep = "\n")
```


## `create_vignette`

Authoring vignettes is a great way to document your package.  I have found
that the parallel development of a 'how to use this package' vignette while
writing and testing the package code has made development and documentation
easier.

Developing the vignette and the code base can be painful.  Consider a
standard .Rmd or .Rnw file with with R code in individual chunks.  Having to
re-evaluate all the chunks can be time consuming and cumbersome depending on
your IDE.  An alternative method is to author `.R` files with the expectation
of using [`knitr::spin`](https://yihui.name/knitr/demo/stitch/) to generate
the `.Rmd` file that then gets processed into the vignette(s).

The `create_vignette` places a template vignette file and makefile into a
`vignettes` directory.#' 

```{r }
create_vignette(name = "egVign.R", path = pkg_dir)
tree(pkg_dir)
```


It is critically important that you **do not use**
`devtools::build_vignettes()` as this will over write the .R file.  If you
need to build the vignettes without building the whole package use the
makefile in the package root directory and call

```
make .vignettes.Rout
```


# Examples:

The best example for this style of package development is the
[qwraps2](https://github.com/dewittpe/qwraps2) package on github.



# Session Info


```{r }
print(sessionInfo(), local = FALSE)
```

