---
title: "Formatted Summary Statistics and Data Summary Tables with qwraps2"
author: "Peter DeWitt"
date: "`r Sys.Date()`"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: true
vignette: >
 %\VignetteIndexEntry{summary-statistics}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---


```{r label = "setup", include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```



# Introduction

It is common for a manuscript to require a data summary table.  The table might
include simple summary statistics for the whole sample and for subgroups.
There are several tools available to build such tables.  In my opinion, though,
most of those tools have nuances imposed by the creators/authors such that other
users need not only understand the tool, but also think like the authors.
I wrote this package to be as flexible and general as possible.  I hope you like
these tools and will be able to use them in your work.

This vignette presents the use of the `summary_table`, `qsummary`, and
`qable` functions for quickly building data summary tables.  These functions
implicitly use the `mean_sd`, `median_iqr`, and `n_perc0` functions from
`qwraps2` as well.

## Prerequisites Example Data Set

We will use a modified version of the `mtcars` data set for examples throughout
this vignette. The following packages are required to run the code in this
vignette and to construct the `mtcars2` `data.frame`.

The `mtcars2` data frame will have three versions of the `cyl` vector: the
original numeric values in `cyl`, a `character` version, and a `factor` version.

```{r }
set.seed(42)
library(magrittr)
library(qwraps2)

# define the markup language we are working in.
# options(qwraps2_markup = "latex") is also supported.
options(qwraps2_markup = "markdown")

data(mtcars)

mtcars2 <-
  dplyr::mutate(mtcars,
                cyl_factor = factor(cyl,
                                    levels = c(6, 4, 8),
                                    labels = paste(c(6, 4, 8), "cylinders")),
                cyl_character = paste(cyl, "cylinders"))

str(mtcars2)
```


Notice that the construction of the `cyl_factor` and `cyl_character` vectors
was done such that the coercion of `cyl_character` to a `factor` will not be the
same as the `cyl_factor` vector; the levels are in a different order.


```{r }
with(mtcars2, table(cyl_factor, cyl_character))
with(mtcars2, all.equal(factor(cyl_character), cyl_factor))
```






# Review of Summary Statistic Functions and Formatting


## Means and Standard Deviations

`mean_sd` will return the (arithmetic) mean and standard deviation for numeric
vector. For example, `mean_sd(mtcars2$mpg)` will return the formatted string.


```{r }
mean_sd(mtcars2$mpg)
mean_sd(mtcars2$mpg, denote_sd = "paren")
```


The default setting for `mean_sd` is to return the mean &plusmn; sd.  In a
table this default is helpful because the default table formatting for counts
and percentages is n (%).

`mean_sd` and other functions are helpful for in-line text too:

    The `r nrow(mtcars2)` vehicles in the `mtcars` data set had an average fuel
    economy of `r mean_sd(mtcars$mpg)` miles per gallon.

produces

> The `r nrow(mtcars2)` vehicles in the `mtcars` data set had an average fuel
> economy of `r mean_sd(mtcars$mpg)` miles per gallon.

## Mean and Confidence intervals
If you need the mean and a confidence interval there is `mean_ci`.
`mean_ci` returns a `qwraps2_mean_ci` object which is a
named vector with the mean, lower confidence limit, and the upper confidence
limit.   The printing method for `qwraps2_mean_ci` objects is a call to the
`frmtci` function.  You an modify the formatting of printed result by adjusting
the arguments pasted to `frmtci`.

```{r }
mci <- mean_ci(mtcars2$mpg)
mci
print(mci, show_level = TRUE)
```




## Median and Inner Quartile Range

Similar to the `mean_sd` function, the `median_iqr` returns the median and the
inner quartile range (IQR) of a data vector.

```{r }
median_iqr(mtcars2$mpg)
```




## Count and Percentages

The `n_perc` function is the workhorse, but `n_perc0` is also provided for ease
of use in the same way that base R has `paste` and `paste0`.  `n_perc` returns
the n (%) with the percentage sign in the string, `n_perc0` omits the
percentage sign from the string.  The latter is good for tables, the former for
in-line text.


```{r }
n_perc(mtcars2$cyl == 4)
n_perc0(mtcars2$cyl == 4)

n_perc(mtcars2$cyl_factor == 4)  # this returns 0 (0.00%)
n_perc(mtcars2$cyl_factor == "4 cylinders")
n_perc(mtcars2$cyl_factor == levels(mtcars2$cyl_factor)[2])

# The count and percentage of 4 or 6 cylinders vehicles in the data set is
n_perc(mtcars2$cyl %in% c(4, 6))
```




## Geometric Means and Standard Deviations

Let $\left\{x_1, x_2, x_3, \ldots, x_n \right\}$ be a sample of size $n$ with
$x_i > 0$ for all $i.$  Then the geometric mean, $\mu_g,$ and geometric standard
deviation are

$$
\begin{equation}
  \mu_g = \left( \prod_{i = 1}^{n} x_i \right)^{\frac{1}{n}} = b^{ \sum_{i =
  1}^{n} \log_{b} x_i },
\end{equation}
$$
and
$$
\begin{equation}
  \sigma_g = b ^ {
  \sqrt{
    \frac{\sum_{i = 1}^{n} \left( \log_{b} \frac{x_i}{\mu_g} \right)^2}{n}
  }
  }
\end{equation}
$$
or, for clarity,
$$
\begin{equation}
  \log_{b} \sigma_g =
  \sqrt{ \frac{\sum_{i = 1}^{n} \left( \log_{b} \frac{x_i}{\mu_g}
  \right)^2}{n}}
\end{equation}
$$

When looking for the geometric standard deviation in R, the simple
`exp(sd(log(x)))` is not exactly correct.  The geometric standard deviation
uses $n,$ the full sample size, in the denominator, where as
the `sd` and `var` functions in R use the denominator $n - 1.$  To get
the geometric standard deviation one should adjust the result by multiplying the
variance by $(n - 1) / n$ or the standard deviation by $\sqrt{(n - 1) / n}.$
See the example below.

```{r }
x <- runif(6, min = 4, max = 70)

# geometric mean
mu_g <- prod(x) ** (1 / length(x))
mu_g
exp(mean(log(x)))
1.2 ** mean(log(x, base = 1.2))

# geometric standard deviation
exp(sd(log(x)))  ## This is wrong

# these equations are correct
sigma_g <- exp(sqrt(sum(log(x / mu_g) ** 2) / length(x)))
sigma_g

exp(sqrt((length(x) - 1) / length(x)) * sd(log(x)))
```


The functions `gmean`, `gvar`, and `gsd` in the
package, provide the geometric
mean, variance, and standard deviation for a sample.

```{r }
gmean(x)
all.equal(gmean(x), mu_g)

gvar(x)
all.equal(gvar(x), sigma_g^2)  # This is supposed to be FALSE
all.equal(gvar(x), exp(log(sigma_g)^2))

gsd(x)
all.equal(gsd(x), sigma_g)
```


`gmean_sd` will provide a quick way for reporting the geometric mean and
geometric standard deviation in the same way that `mean_sd` does for the
arithmetic mean and arithmetic standard deviation:

```{r }
gmean_sd(x)
```





# Building a Data Summary Table

Objective: build a table reporting summary statistics for some of the variables
in the `mtcars2` `data.frame` overall and within subgroups.  We'll start with
something very simple and build up to something bigger.

Let's report the min, max, and mean (sd) for continuous variables and n (%) for
categorical variables.  We will report `mpg`, `disp`, `wt`, and `gear` overall
and by number of cylinders.

The function `summary_table`, along with some `dplyr` functions will do the work
for us.  `summary_table` takes two arguments:

1. `x` a (`grouped_df`) data.frame.
2. `summaries` a list of summaries.  This is a list-of-lists.  The outer list
   defines the row groups and the inner lists define the specif summaries.
   The default is generated by the `qsummary` function.


```{r }
args(summary_table)
```


Let's build a list-of-lists to pass to the `summaries` argument of
`summary_table`.  Additional examples and tools for building the
list-of-lists are given in the following section.  The immediate example is
provided to demonstrate how to use the `summary_table` method.

The inner lists are named `formula`e defining the wanted
summary.  These `formula`e are passed through `dplyr::summarize` to generate
the table.  The names are important, as they are used to label row groups and row
names in the table.  The arguemnt for the functions below use the `.data`
pronoun for tidy evaluation (see `help(topic = ".data", package = "rlang")`).
The use of this pronoun is not mandatory, however, the use of the pronoun is
strongly encouraged.


```{r }
our_summary1 <-
  list("Miles Per Gallon" =
       list("min" = ~ min(.data$mpg),
            "max" = ~ max(.data$mpg),
            "mean (sd)" = ~ qwraps2::mean_sd(.data$mpg)),
       "Displacement" =
       list("min" = ~ min(.data$disp),
            "median" = ~ median(.data$disp),
            "max" = ~ max(.data$disp),
            "mean (sd)" = ~ qwraps2::mean_sd(.data$disp)),
       "Weight (1000 lbs)" =
       list("min" = ~ min(.data$wt),
            "max" = ~ max(.data$wt),
            "mean (sd)" = ~ qwraps2::mean_sd(.data$wt)),
       "Forward Gears" =
       list("Three" = ~ qwraps2::n_perc0(.data$gear == 3),
            "Four"  = ~ qwraps2::n_perc0(.data$gear == 4),
            "Five"  = ~ qwraps2::n_perc0(.data$gear == 5))
       )
```


Building the table is done with a call to `summary_table`:


```{r results = "asis"}
### Overall
whole <- summary_table(mtcars2, our_summary1)
whole
```


The `summary_table` will work with grouped data frames too.

```{r results = "asis"}
### By number of Cylinders
by_cyl <- summary_table(dplyr::group_by(mtcars2, cyl_factor), our_summary1)
by_cyl
```


To report a table with both the whole sample summary and conditional columns
together:

```{r results = "asis"}
both <- cbind(whole, by_cyl)
both
```


If you want to change the column names, do so via the `cnames` argument to
`qable` via the print method for `qwraps2_summary_table` objects.  Any argument
that you want to send to `qable` can be sent there when explicitly using the
`print` method for `qwraps2_summary_table` objects.

```{r results = "asis"}
print(both,
      rtitle = "Summary Statistics",
      cnames = c("Col 0", "Col 1", "Col 2", "Col 3"))
```



## Easy building of the summaries

The task of building the `summaries` list-of-lists can be tedious.  `qsummary`
is designed to make it easier.  `qsummary` will use a set of predefined
functions to summarize numeric columns of a `data.frame`, a set of arguments
to pass to `qwraps2::n_perc` for categorical (`character` and `factors`)
variables.

By default, calling `summary_table` will use the default summary metrics
defined by `qsummary`.  The purpose of `qsummary` is to provide the same
summary for all numeric variables within a data.frame and a single style of
summary for categorical variables within the data.frame.  For example, the
default summary for a set of variables from the the `mtcars2` data set is

```{r }
mtcars2 %>%
  dplyr::select(.data$mpg, .data$cyl_factor, .data$wt) %>%
  qsummary(.)
```


That default summary is used for a table as follows:

```{r label="summary_table_mtcars2_default", results = "asis"}
mtcars2 %>%
  dplyr::select(.data$mpg, .data$cyl_factor, .data$wt) %>%
  summary_table(.)
```


Now, say we want to only report the minimum and maximum for each of the
numeric variables and for the categorical variables we want two show the
denominator for each category and for the percentage, to one digit with the
percent symbol in the table.
Note that when defining the list of numeric_summaries that the argument place
holder is the `%s` character.

```{r }
new_summary <-
  mtcars2 %>%
  dplyr::select(.data$mpg, .data$cyl_factor, .data$wt) %>%
  qsummary(.,
           numeric_summaries = list("Minimum" = "~ min(%s)",
                                    "Maximum" = "~ max(%s)"),
           n_perc_args = list(digits = 1, show_symbol = TRUE, show_denom = "always"))
```


The resulting table is:

```{r results = "asis"}
summary_table(mtcars2, new_summary)
```


The summary can easily be used on a grouped `data.frame`.

```{r results = "asis"}
mtcars2 %>%
  dplyr::group_by(.data$am) %>%
  summary_table(., new_summary)
```






## Adding P-values to a Summary Table

There are many, many different ways to format data summary tables. Adding
p-values to a table is just one thing that can be done in more than one way.
For example, if a row group reports the counts and percentages for each level
of a categorical variable across multiple (column) groups, then I would argue
that the p-value resulting from a chi square test or a Fisher exact test
would be best placed on the line of the table labeling the row group.
However, say we reported the minimum, median, mean, and maximum with in a
row group for one variable.  The p-value from a t-test, or other meaningful
test, for the difference in mean I would suggest should be reported on the
line of the summary table for the mean, not the row group itself.

With so many possibilities I have reserved construction of a p-value column
to be ad hoc.  Perhaps an additional column wouldn't be used and the p-values
are edited into row group labels, for example.

If you want to add a p-value column to a `qwraps2_summary_table` object you
can with some degree of ease.  Note that `qwraps2_summary_table` objects are
just character matrices.

```{r }
both %>% str
```


Let's added p-values for testing the difference in the mean between the three
cylinder groups.

```{r }
# difference in means
mpvals <-
  list(lm(mpg ~ cyl_factor,  data = mtcars2),
       lm(disp ~ cyl_factor, data = mtcars2),
       lm(wt ~ cyl_factor,   data = mtcars2)) %>%
  lapply(aov) %>%
  lapply(summary) %>%
  lapply(function(x) x[[1]][["Pr(>F)"]][1]) %>%
  lapply(frmtp) %>%
  do.call(c, .)

# Fisher test
fpval <- frmtp(fisher.test(table(mtcars2$gear, mtcars2$cyl_factor))$p.value)
```


Adding the p-value column is done as follows:

```{r }
both <- cbind(both, "P-value" = "")
both[grepl("mean \\(sd\\)", rownames(both)), "P-value"] <- mpvals
a <- capture.output(print(both))
a[grepl("Forward Gears", a)] %<>% sub("&nbsp;&nbsp;\\ \\|$", paste(fpval, "|"), .)
```


and the resulting table is:

```{r results = "asis"}
cat(a, sep = "\n")
```


Another option you might consider is to have the p-value in the row group
name.  Consider the following construction.  The p-values are added to the
names of the row groups when building the summary table.

```{r results = "asis"}
gear_summary <-
  list("Forward Gears" =
       list("Three" = ~ qwraps2::n_perc0(.data$gear == 3),
            "Four"  = ~ qwraps2::n_perc0(.data$gear == 4),
            "Five"  = ~ qwraps2::n_perc0(.data$gear == 5)),
       "Transmission" =
       list("Automatic" = ~ qwraps2::n_perc0(.data$am == 0),
            "Manual"  = ~ qwraps2::n_perc0(.data$am == 1))
       ) %>%
setNames(.,
         c(
         paste("Forward Gears: ", frmtp(fisher.test(xtabs( ~ gear + cyl_factor, data = mtcars2))$p.value)),
         paste("Transmission: ",  frmtp(fisher.test(xtabs( ~ am + cyl_factor, data = mtcars2))$p.value)))
         )

summary_table(dplyr::group_by(mtcars2, cyl_factor), gear_summary)
```





## Using Variable Labels

Some data management paradigms will use attributes to keep a label associated
with a variable in a data.frame.  Notable examples are the
[Hmisc](https://cran.r-project.org/package=Hmisc) and
[sjPlot](https://cran.r-project.org/package=sjPlot).  If you associate a
label with a variable in the data frame the that label will be used when
building a summary table.  This feature was suggested
https://github.com/dewittpe/qwraps2/issues/74 and implemented thusly:

```{r }
new_data_frame <-
  data.frame(age = c(18, 20, 24, 17, 43),
             edu = c(1, 3, 1, 5, 2),
             rt  = c(0.01, 0.04, 0.02, 0.10, 0.06))

# Set a label for the variables
attr(new_data_frame$age, "label") <- "Age in years"
attr(new_data_frame$rt,  "label") <- "Reaction time"

# mistakenly set the attribute to name instead of label
attr(new_data_frame$edu, "name") <- "Education"
```


When calling `qsummary` the provide labels for the age and rt variables will
be used.  Since the attribute "label" does not exist for the edu variable,
edu will be used in the output.

```{r }
qsummary(new_data_frame)
```


This behavior is also seen with the `summary_table` call

```{r results = "asis"}
summary_table(new_data_frame)
```





# Session Info


```{r }
print(sessionInfo(), local = FALSE)
```

